Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index

### 2016-06-14 swsc.tex
Updates from ShareLaTeX

The quantities $l$ in the RBF, $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}. Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The method of assigning values to the \emph{hyper-parameter} is crucial in the model building process (see section \ref{sec:hyp} below). In this study, we construct Gaussian Process regression models with linear polynomial kernels for \emph{one step ahead} prediction of the $Dst$ index.

\subsection{Optimizing Hyper-parameters} \label{sec:hyp}

When applying \emph{Gaussian Process} models to data, one must choose appropriate values to the kernel hyper-parameters and noise level ($\sigma^2$). This is known as the problem of \emph{model selection}.

Common examples of criteria used for \emph{model selection} include data likelihood, cross validation performance scores and validation set performance scores among others. Since model selection criteria are non-convex and non-smooth in general, gradient free approaches are used to find approximate values of local optima.

The simplest way to perform \emph{model selection} is called \emph{grid search} which entails evaluating the above criteria on a fixed grid which is constructed by the tensorial product of grids constructed on each hyper-parameter. There are more sophisticated methods for \emph{model tuning} such as \emph{coupled simulated annealing} (\citet{Xavier-De-Souza2010}) and \emph{simplex search} (\citet{Nelder1965}) but they are computationally more expensive.

In this study we employ \emph{grid search} to find the best performing set of hyper-parameter assignments inside a predefined grid of possible hyper-parameter values.

@@

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{energyLandscape.png}
      \caption{Contour plot of model selection metric (\emph{RMSE}) for \emph{GP-ARX} over the validation data.}
         \label{fig:energy}
   \end{figure}

@@

Model selection of \emph{GP-AR} and \emph{GP-ARX} is performed by grid search methodology and the root mean square error (\emph{RMSE}) on the validation set is used to select the values of the model hyper-parameters. Figure \ref{fig:energy} shows a contour plot of the \emph{RMSE} computed on the validation data for the \emph{GP-ARX} model, the x axis represents the intercept $b$ of the polynomial kernel while the y axis represents $\sigma$ the standard deviation of the noise of the GP model as seen in section \ref{sec:inference}. In the experiments performed the best performing values of the hyper-parameters outputted by the grid search routing are $b = 0.002,\ \sigma = 1.994$.

>>>>

The quantities $l$ in the RBF, $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}. Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The method of assigning values to the \emph{hyper-parameter} is crucial in the model building process.

In this study, we construct Gaussian Process regression models with linear kernels as shown in equation \ref{eq:usedKernel}, for \emph{one step ahead} prediction of the $Dst$ index. Our choice of kernel leaves us with only one adjustable hyper-parameter, the model noise $\sigma$.

\begin{equation}\label{eq:usedKernel}
    K(\mathbf{x}, \mathbf{y}) = \mathbf{x}^\intercal \mathbf{y}
\end{equation}

We initialize a grid of values for $\sigma$ the model noise and use the error on a predefined validation data set to choose the best performing value of $\sigma$. While constructing the grid of possible values for $\sigma$ it must be ensured that $\sigma > 0$ so that the kernel matrix constructed on the training data is non-singular.

@@

@@

Model selection of \emph{GP-AR} and \emph{GP-ARX} is performed by grid search methodology and the root mean square error (\emph{RMSE}) on the validation set is used to select the values of the model hyper-parameters. In the experiments performed the best performing values of the hyper-parameters outputted by the grid search routine are $\sigma = 0.45$ for the \emph{GP-ARX} and $\sigma = 1.173$ for \emph{GP-AR}.
